Direct Entropy Measurement in a Mesoscopic Quantum System

### 2018-02-23 paper/draft.tex
text updates

%%%%%%% intro material %%%%%%%%%
\textbf{The thermodynamic properties of electronic systems offer important insights into the nature of their ground states, and can probe exotic quasiparticles that may emerge due to interactions or non-trivial topology.  Systems that are difficult to clearly identify through standard conductance measurements may be characterized more precisely with a thermodynamic measurement. For example, the purported non-Abelian exchange statistics of Moore-Read quasiparticles in the $\nu = \frac{5}{2}$ fractional quantum Hall state, or of Majorana quasiparticles in a topological superconductor, could be clearly distinguished from the Abelian case by an entropy measurement \cite{Cooper2009, Smirnov2015}.  Such measurements are typically based on bulk properties, like heat capacity, that are straightforward to observe in macroscopic samples but difficult to access in mesoscopic systems that may consist of only a few electrons.  Here, we develop a mesoscopic circuit to make a direct entropy measurement on the first, second, and third electron occupations in a GaAs quantum dot.  This system is chosen for having well-known entropy based on simple spin states\cite{Tarucha1996, Ciorga2000, Duncan2000, Lindemann2002, Potok2003, Hofmann2016}.  The precision of this technique, quantifying the entropy of a single spin-$\frac{1}{2}$ to within 5\% of the expected value ($k_B \ln{2}$), shows its future potential for probing more exotic systems, from Majorana states to non-Fermi liquid Kondo states, that are clearly identified by a residual low-temperature entropy\cite{Alkurtass2016}.}

@@

Our approach is analogous to the milestone of spin-to-charge conversion achieved over a decade ago, by which unmeasurably-small magnetic moments of a single spin were detected indirectly by transforming into the presence or absence of an electron charge \cite{Elzerman2004, Ono2004}.  Instead, we perform an entropy-to-charge conversion, making use of the Maxwell relation
%
\begin{align}
\label{eqn:max}
        \left(\frac{\partial \mu}{\partial T}\right)_{p,N} &= -\left(\frac{\partial S}{\partial N}\right)_{p,T}
\end{align}
%
that connects changes in entropy, particle number, and temperature ($S$, $N$, and $T$, respectively) to changes in the chemical potential, $\mu$, a quantity that is simple to measure and control. The fixed pressure condition of Eq.~\ref{eqn:max} is met by working well below the Fermi temperature of the 2DEG, $T_F \sim \SI{5}{\kelvin}$, where degeneracy pressure dominates \cite{Landau1969}.

The Maxwell relation in Eq.~\ref{eqn:max} forms the basis of two theoretical proposals to measure non-Abelian exchange of quasiparticles in the the $\nu = \frac{5}{2}$ state, by detecting changes in the quasiparticle chemical potential with temperature, which are proportional to the entropy derivative, $\frac{\partial S}{\partial N}$ \cite{Cooper2009,Ben-Shach2013}. Applying the language of quantum dots to Eq.~\ref{eqn:max}, the entropy difference between the $N-1$ to $N$ quantum dot ground states ($\Delta S_{N-1\rightarrow N}$ for $\Delta N=1$) is measured through the shift with temperature in the electrochemical potential, $\mu_N$, needed to add the $N$th electron to the dot; the gate voltage at which the $N-1$ to $N$ transition is observed shifts linearly with $\mu_{N}$.

The mesoscopic circuit shown in Fig.~\ref{fig:fig1}a uses electrostatic gates to realize an electron reservoir (central region; hereafter ``reservoir") in thermal and diffusive equilibrium with a few-electron quantum dot (``dot") coupled to its right side.  The reservoir temperature, $T$, could be rapidly (up to at least \SI{100}{\hertz}) increased above the mixing chamber temperature, $T_{MC}$, by Joule heating through the QPC (``heater") on the left side.  The occupation of the dot was measured using an adjacent quantum point contact (``charge sensor")\cite{Staring2007, Thierschmann2015} [change these references/add additional].  Applying a more positive voltage to the gate adds the $N$th electron to the dot when $\mu_{N}$ drops below the Fermi level of the reservoir, $E_F$. Figure 1b illustrates such a transition \textemdash a step in the charge sensor conductance \textemdash thermally broadened by the reservoir temperature.  The gate voltage corresponding to the midpoint of the transition, $V_{mid}$, marks the electrochemical potential at which the probabilities of finding $N-1$ or $N$ electrons on the dot are equal.  When $\mu_N$ shifts with temperature, $V_{mid}$ also shifts (Fig.~1c); it is this signature that forms the basis of our experiment.

The relation between entropy and any temperature-induced shift in $\mu_{N}$ with respect to $E_F$ can also be understood in the language of detailed balance.  The $N$th electron tunnels back and forth between an $N-1$-electron dot and reservoir when there are available states in both the dot and reservoir, that is, when $\mu_{N}$ is within thermal broadening of $E_F$.  The tunnel rates into and out of the dot, $\Gamma_{in}=\Gamma_{N-1\rightarrow N}$ and $\Gamma_{out}=\Gamma_{N\rightarrow N-1}$, depend on degeneracy of the $N-1$ and $N$ ground states. Consider, for example, the first electron tunneling into an empty dot: both spin-up and -down states are available.  Tunneling out, the electron (whether spin-up or -down) must tunnel into reservoir states with the same spin, that is, only half of the empty reservoir states are available.  As a result, $\Gamma_{in} = 2\Gamma_{out}$ when $\mu_{N}=E_F$.  Extending this logic to a general case gives the result derived from detailed balance \cite{Gustavsson2009, Beenakker1991}, 
%
\begin{align}
	\Gamma_{in} &=  g_{N} \Gamma f(E_F - \mu_{N}) \nonumber \\
	\Gamma_{out} &= g_{N-1} \Gamma [1 - f(E_F - \mu_{N})] \label{eqn:rates}
\end{align}
%
where $\Gamma$ is a coupling constant, $f$ is the Fermi function, and $g_{N(N-1)}$ is the degeneracy of the $N(N-1)$ electron state.

The equal probability condition for $N-1/N$ at $V_{mid}$ corresponds to  incoming and outgoing rates being equal.  From Eq.~\ref{eqn:rates}$, \Gamma_{in} = \Gamma_{out}$ corresponds to $\mu_{N} = E_F$ as long as $g_{N-1}=g_{N}$, but to $\mu_{N} > E_F$ when $g_{N-1} > g_{N}$ and vice versa.   The ratio $g_{N}/g_{N-1}$ is given by the shift in $V_{mid}$ with temperature, giving $\Delta S_{N-1\rightarrow N}$ via the Boltzmann entropy, $S_{N}=k_{B} \ln{g_N}$. Previous works have explored the relationship between tunnel rates and degeneracy using time-resolved transport spectroscopy and coupling of quantum dots to atomic force microscope oscillations \cite{Cockins2010, Bennett2010, Beckel2014, Hofmann2016}. The approach presented here illustrates the thermodynamic equivalent of the rate measurement, and extends entropy measurements to a wide set of applications where tunneling processes may not be possible to observe in real-time.

@@

To proceed with the entropy measurement, the quantum dot (Fig.~\ref{fig:fig1}a) was tuned such that the source was weakly tunnel coupled to the reservoir with the drain closed.   The conductance of the charge sensor was tuned to $g_{sens}{\sim}e^2/h$ where it was most sensitive to charge on the dot.  The addition of the first electron to the dot was marked by a decrease in $g_{sens}$ that is well fit by the expected form for a thermally-broadened transition:
%
\begin{align}
\label{eqn:g}
        g_{sens}(V,\Theta) &= G_0 \tanh\left(\frac{V-V_{mid}(\Theta)}{2\Theta}\right)  \\
                        &\quad + G_1\left[V-V_{mid}(\Theta)\right] + G_2 \nonumber
\end{align}
%
where $G_0$ quantifies the sensor sensitivity, $\Theta = \frac{k_B T}{\alpha e}$ is the thermal broadening expressed in units of gate voltage, $\alpha$ is the lever arm $\alpha\equiv\partial \mu_{N}/\partial V_G$, and $G_1$ reflects the cross capacitance between the charge sensor and plunger gate. Fig.~\ref{fig:fig2}a shows two such transition curves with the high temperature curve being thermally broadened by driving a current, $I_{heat}$ through the QPC heater. With no Joule heating through the heater, $\Theta$ followed $T_{MC}$ down to \SI{<100}{\milli\kelvin} (Fig.~\ref{fig:fig2}b), validating the approximation of thermal broadening used throughout this experiment.

The data in Fig.~\ref{fig:fig2}c, and corresponding fits, offer an example of how $\Delta S_{0\rightarrow 1}$ was measured across the $0 \rightarrow 1$ transition.  A comparison of the curves in Fig.~\ref{fig:fig2}a shows that $g_{sens}$ drops as temperature rises immediately before the midpoint of the transition ($V_G-V_{mid}<0$), whereas $g_{sens}$ rises with temperature for $V_G-V_{mid}>0$.   Oscillating the reservoir temperature by $\delta T$ using an AC current through the heater while measuring oscillations in the charge sensor signal, $\delta g_{sens}$, with a lock in amplifier (see Methods) yields the characteristic peak-dip structure seen in Fig.~\ref{fig:fig2}c.  The expected lineshape of such a curve is given by the derivative of Eq.~\ref{eqn:g} with respect to temperature, expressed as $\delta g_{sens}$ due to a temperature change $\delta T$:
%
\begin{align}
\label{eqn:dg}
        \delta g_{sens}(V_G, \Theta) &\propto -\delta T \left[ \frac{V_G-V_{mid}(\Theta)}{2\Theta} + \frac{\epsilon}{2} \right]\times \\
        				      &\quad\cosh^{-2}\left(\frac{V_G-V_{mid}(\Theta)}{2\Theta}\right) + const. \nonumber
\end{align}
%
where $\epsilon=\frac{\partial V_{mid}}{\partial \Theta}$. Using the lever arm to convert $V_{mid}$ to electrochemical potential, $\epsilon$ is simply the change in entropy (scaled by $k_B$) through Eq.~\ref{eqn:max}:
%
\begin{align}
\label{eqn:epsilon}
        \epsilon &= \frac{\partial V_{mid}}{\partial \Theta} = 
        \frac{1}{k_B} \frac{\partial \mu}{\partial T} = 
        -\frac{1}{k_B} \Delta S_{N-1\rightarrow N}
\end{align}
%

From Eqs.~\ref{eqn:dg} and \ref{eqn:epsilon}, $\delta g_{sens}(V_G)$ would be perfectly antisymmetric around $V_{mid}$ if $\Delta S_{N-1\rightarrow N}$ were zero, whereas the data in Fig.~\ref{fig:fig2}c are clearly not antisymmetric. A best fit to Eqs.~\ref{eqn:dg} and \ref{eqn:epsilon} yields $\Delta S_{0\rightarrow 1}=0.72 \pm 0.01 k_B$, close to what would be expected for the $0 \rightarrow 1$ electron transition at zero magnetic field: $S_0$ for the empty state should be $k_B \ln{1}=0$ whereas for the spin degenerate one-electron state one expects $S_1=k_B\ln{2}$, giving $\Delta S_{0\rightarrow 1}\equiv S_1 - S_0 =k_B\ln{2}$.

Fig.~\ref{fig:fig2}e shows that $\Delta S$,  determined from fits to data analogous to Fig.~\ref{fig:fig2}c, remains constant over a range of $\delta T$ (I$_{heat}$ AC).  To confirm the measurement lies within the the relevant limits, the magnitude of the temperature change, $\delta T$ is calibrated in Fig.~\ref{fig:fig2}d and the corresponding shift in $V_{mid}$ is plotted long the right side of Fig.~\ref{fig:fig2}e. However, it is important to note that the change in entropy, $\Delta S$, is determined from the fit parameter, $\epsilon$, with no further calibration.

%%% figure 3 %%%
\begin{figure}
        \includegraphics[width=1.0\columnwidth]{../figures/figure_3.pdf}
        \caption{\label{fig:fig3}(a) Changes in entropy for $N=0 \rightarrow 1$ and $2 \rightarrow 3$ transitions, overlaid to highlight similar behaviors.  Each datapoint corresponds to a fit to a single $\delta g_{sens}(V_G)$ trace; multiple scans are carried out at various parallel magnetic fields.   (b) and (c) Characteristic $\delta g_{sens}$ traces from which the data in (a) were extracted. The two data points corresponding to (b) and (c) are shown as large markers in (a). (e) Bias spectroscopy data for the $N=0 \rightarrow 1$ transition. Dashed line at $V_{SD}$ = \SI{700}{\micro\electronvolt} shows where data in (f) are taken. (f) Fixed bias data showing fits to Zeeman splitting of the ground state (dashed lines) from which $g = -0.42 \pm 0.01$ is extracted.}
\end{figure}

Confirmation that the measured $\Delta S$ derives from spin degeneracy can be seen in its evolution with in-plane magnetic field, $B_\parallel$. Figure \ref{fig:fig3}a compares $\Delta S(B_\parallel)$ for the 0-1 and 2-3 electron transitions.  The 2-electron ground state is a spin singlet, while the 3-electron ground state adds an additional unpaired electron, so both $\Delta S_{0 \rightarrow 1}$ and $\Delta S_{2 \rightarrow 3}$ are expected to correspond to transitions from total spin zero to total spin one-half. The entropies of the 1- and 3-electron states are suppressed when Zeeman splitting lifts the spin degeneracy of the unpaired electron. This behavior is described by the Gibbs entropy for a two-level system:
%
\begin{align}
\label{eqn:gibbs}
        S &= k_B \sum_{i=\pm} p_{i}(B_\parallel, T) \ln{ p_{i}(B_\parallel,T) }
\end{align}
%
where $p_{\pm}(B_\parallel, T) = (1+ e^{\mp \frac{g\mu_B B_{\parallel}}{k_B T}})^{-1}$ are the probabilities for the unpaired electron to be in the spin up/down states at a given field and temperature. Fits to Eq.~\ref{eqn:gibbs} are shown in Fig.~\ref{fig:fig3}a for both transitions, with the ration $g/T$ and an added scaling $\Delta S(B=0)$ as free parameters.

The entropy measurements yield $\Delta S_{0 \rightarrow 1}=(0.93 \pm 0.02) k_B \ln{2}$ and $\Delta S_{2 \rightarrow 3}=(0.98 \pm 0.01) k_B \ln{2}$ at zero magnetic field, collapsing to zero at high field where spin degeneracy is broken.  At a qualitative level, Figs.~\ref{fig:fig3}b and c show $\delta g_{sens}$ crossover from asymmetric to antisymmetric as $\Delta S$ goes to zero at high field. Choosing $T$ to be midway between the low ($T_{MC}$) and high ($T_{MC}+\delta T_{RMS}$) points of the temperature oscillation for each data set, yields $g=-0.39 \pm 0.01 $ and $g=-0.45 \pm 0.01 $ for the $0\rightarrow 1$ and $2\rightarrow 3$ transitions, respectively. These are consistent with previous few electron dot measurements and the value measured separately in Fig.~\ref{fig:fig3}e using bias spectroscopy \cite{Cronenwett1998,Hanson2003,Zumbuhl2004}.

%%% figure 4 %%%
\begin{figure}
        \includegraphics[width=1.0\columnwidth]{../figures/figure_4.pdf}
        \caption{\label{fig:fig4}(a) Change in entropy, extracted from $\delta g_{sens}$ fits at varying parallel field. Dashed line shows fit to Eq.~\ref{eqn:gibbs}. Values stated for $\Delta S$ are with respect to the vertical offset apparent in the data. (b), (c), and (d) show characteristic $\delta g_{sens}$ traces from which the data in (a) were extracted. These data points are show as large markers in (a). (e) Bias spectroscopy data for the $N=1 \rightarrow 2$ transition. Transitions to the 2-electron triplet state correspond to the lines appearing at $V_{SD} = \SI{\pm 320}{\micro\electronvolt}$. Dashed line at $V_{SD}$ = \SI{1250}{\micro\electronvolt} shows where data in (f) are taken. (f) Fixed bias data in parallel field. Triplet level is split into $T_+$ and $T_0$ levels with a third $T_{-}$ level not visible here. At \SI[input-protect-tokens]{9}{\tesla} $T_+$ becomes degenerate with $S$. $g=-0.40\pm0.04$ is determined using $T_0$ and $T_+$ fits (dashed).}
\end{figure}

The $1\rightarrow 2$ transition (Fig.~\ref{fig:fig4}a) can be understood as the inverse of Fig \ref{fig:fig3}a for $B_\parallel < \SI{5}{\tesla}$.  There, the 2-electron ground state remains a spin singlet ($S_2=0$) while the 1-electron ground state entropy goes from $k_B\ln{2}$ to zero due to Zeeman splitting. At higher field, the 1-electron ground state remains non-degenerate ($S_1=0$) while the 2-electron ground state becomes two-fold degenerate where the spin singlet ($S$) and triplet ($T_+$) states cross. The singlet-triplet crossing is seen clearly around at \SI{9}{\tesla} in bias spectroscopy data for the $1\rightarrow 2$ transition (Fig.~\ref{fig:fig4}e).

>>>>

%%%%%%% intro material %%%%%%%%%
\textbf{The thermodynamic properties of electronic systems offer important insights into the nature of their ground states, and can probe exotic quasiparticles that may emerge due to interactions or non-trivial topology.  Systems that are difficult to clearly identify through standard conductance measurements may be characterized more precisely with a thermodynamic measurement. For example, the purported non-Abelian exchange statistics of Moore-Read quasiparticles in the $\nu = \frac{5}{2}$ fractional quantum Hall state, or of Majorana quasiparticles in a topological superconductor, could be clearly distinguished from the Abelian case by an entropy measurement \cite{Cooper2009, Smirnov2015}.  Such measurements are typically based on bulk properties, like heat capacity, that are clearly observed in macroscopic samples but are immeasurably small in systems that may consist of only a few electrons.  Here, we develop a mesoscopic circuit to make direct entropy measurements of the first, second, and third electron ground states in a GaAs quantum dot.  This system is chosen for having well-known entropy based on simple spin states\cite{Tarucha1996, Ciorga2000, Duncan2000, Lindemann2002, Potok2003, Hofmann2016}.  The precision of this technique, quantifying the entropy of a single spin-$\frac{1}{2}$ to within 5\% of the expected value ($k_B \ln{2}$), shows its future potential for probing more exotic systems, from Majorana states to non-Fermi liquid Kondo states, that are clearly identified by their residual low-temperature entropy\cite{Alkurtass2016}.}

@@

Our approach is analogous to the milestone of spin-to-charge conversion achieved over a decade ago, in which the infinitesimal magnetic moments of a single spin were detected by transforming them into the presence or absence of an electron charge \cite{Elzerman2004, Ono2004}.  Following this logic, we perform an entropy-to-charge conversion, making use of the Maxwell relation
%
\begin{align}
\label{eqn:max}
        \left(\frac{\partial \mu}{\partial T}\right)_{p,N} &= -\left(\frac{\partial S}{\partial N}\right)_{p,T}
\end{align}
%
that connects changes in entropy, particle number, and temperature ($S$, $N$, and $T$, respectively) to changes in the chemical potential, $\mu$, a quantity that is simple to measure and control. The fixed pressure condition of Eq.~\ref{eqn:max} is met by working well below the Fermi temperature of the 2DEG, $T_F \sim \SI{5}{\kelvin}$, where degeneracy pressure dominates \cite{Landau1969}.

The Maxwell relation in Eq.~\ref{eqn:max} forms the basis of two theoretical proposals to measure non-Abelian exchange of quasiparticles in the the $\nu = \frac{5}{2}$ state, by detecting a change in quasiparticle chemical potential with temperature, which is proportional to the entropy derivative, $\frac{\partial S}{\partial N}$ \cite{Cooper2009,Ben-Shach2013}. Applying the language of quantum dots to Eq.~\ref{eqn:max}, the entropy difference between the $N-1$ to $N$ quantum dot ground states ($\Delta S_{N-1\rightarrow N}$ for $\Delta N=1$) is measured through the shift with temperature in the electrochemical potential, $\mu_N$, needed to add the $N$th electron to the dot; the gate voltage at which the $N-1$ to $N$ transition is observed is  proportional to $\mu_{N}$.

The mesoscopic circuit shown in Fig.~\ref{fig:fig1}a uses electrostatic gates to realize an electron reservoir (central region; hereafter ``reservoir") in thermal and diffusive equilibrium with a few-electron quantum dot (``dot") coupled to its right side.  The reservoir temperature, $T$, could be rapidly (up to at least \SI{100}{\hertz}) increased above the mixing chamber temperature, $T_{MC}$, by Joule heating through the QPC (``heater") on the left side.  The occupation of the dot was measured using an adjacent quantum point contact (``charge sensor") \cite{Field1993, Staring2007, Thierschmann2015} [change these references/add additional].  Applying a more positive voltage to the gate adds the $N$th electron to the dot when $\mu_{N}$ drops below the Fermi level of the reservoir, $E_F$. Figure~\ref{fig:fig1}b illustrates such a transition \textemdash seen as a step in the charge sensor conductance \textemdash that is thermally broadened by the reservoir temperature.  The gate voltage corresponding to the midpoint of the transition, $V_{mid}$, marks the electrochemical potential at which the probabilities of finding $N-1$ or $N$ electrons on the dot are equal.  When $\mu_N$ shifts with temperature, $V_{mid}$ also shifts (Fig.~\ref{fig:fig1}c); it is the measurement of this signature that forms the basis of our experiment.

The relation between entropy and any temperature-induced shift in $\mu_{N}$ with respect to $E_F$ can be understood through detailed balance.  The $N$th electron tunnels back and forth between an $N-1$-electron dot and reservoir when there are available states in both the dot and reservoir, that is, when $\mu_{N}$ is within thermal broadening ($\sim k_B T$) of $E_F$.  The tunnel rates into and out of the dot, $\Gamma_{in}=\Gamma_{N-1\rightarrow N}$ and $\Gamma_{out}=\Gamma_{N\rightarrow N-1}$, depend on degeneracy of the $N-1$ and $N$ ground states. Consider, for example, the first electron tunneling into an empty dot: both spin-up and -down states are available on the dot.  Tunneling out, the electron (whether spin-up or -down) must tunnel into reservoir states with the same spin, that is, only half of the empty reservoir states are available.  As a result, $\Gamma_{in} = 2\Gamma_{out}$ when $\mu_{N}=E_F$.  Extending this logic to a general case gives the result derived from detailed balance \cite{Beenakker1991, Gustavsson2009}, 
%
\begin{align}
	\Gamma_{in} &=  g_{N} \Gamma f(E_F - \mu_{N}) \nonumber \\
	\Gamma_{out} &= g_{N-1} \Gamma [1 - f(E_F - \mu_{N})] \label{eqn:rates}
\end{align}
%
where $\Gamma$ is a coupling constant, $f$ is the Fermi function, and $g_{N(N-1)}$ is the degeneracy of the $N(N-1)$ electron state.

The equal probability condition for $N-1$- and $N$-electrons at $V_{mid}$ corresponds to the incoming and outgoing rates being equal. From Eq.~\ref{eqn:rates}$, \Gamma_{in} = \Gamma_{out}$ corresponds to $\mu_{N} = E_F$ only in the case $g_{N-1}=g_{N}$, but to $\mu_{N} > E_F$ when $g_{N-1} > g_{N}$ and to $\mu_{N} < E_F$ when $g_{N-1} < g_{N}$. The ratio $g_{N}/g_{N-1}$ is given by the shift in $V_{mid}$ ($\propto \mu_{N}$) with temperature, giving $\Delta S_{N-1\rightarrow N}$ via the Boltzmann entropy, $S_{N}=k_{B} \ln{g_N}$. Previous works have explored the relationship between tunnel rates and degeneracy using time-resolved transport spectroscopy and coupling of quantum dots to atomic force microscope oscillations \cite{Cockins2010, Bennett2010, Beckel2014, Hofmann2016}. The approach presented here illustrates the thermodynamic equivalent of the rate measurement, and extends entropy measurements to a wide set of applications where tunneling processes may not be possible to observe in real-time.

%%%%%%%%%%

@@

To proceed with the entropy measurement, the quantum dot was tuned such that the source was weakly tunnel coupled to the reservoir with the drain closed.  The conductance of the charge sensor was tuned to $g_{sens}{\sim}e^2/h$, where it was most sensitive to charge on the dot.  The addition of the first electron to the dot was marked by a decrease in $g_{sens}$ that is well fit by the expected form for a thermally-broadened transition:
%
\begin{align}
\label{eqn:g}
        g_{sens}(V,\Theta) &= G_0 \tanh\left(\frac{V-V_{mid}(\Theta)}{2\Theta}\right)  \\
                        &\quad + G_1\left[V-V_{mid}(\Theta)\right] + G_2 \nonumber
\end{align}
%
where $G_0$ represents the sensor sensitivity, $\Theta = \frac{k_B T}{\alpha e}$ is the thermal broadening (expressed in units of gate voltage), $\alpha\equiv d \mu_{N}/d V_G$ is the lever arm, and $G_1$ reflects the cross capacitance between the charge sensor and plunger gate. Fig.~\ref{fig:fig2}a shows two such transition curves with the high temperature curve being thermally broadened by driving a current, $I_{heat}$ through the QPC heater. With no Joule heating through the heater, $\Theta$ followed $T_{MC}$ down to \SI{<100}{\milli\kelvin} (Fig.~\ref{fig:fig2}b), validating the approximation of thermal broadening used throughout this experiment.

The data in Fig.~\ref{fig:fig2}c, and corresponding fits, offer an example of how $\Delta S_{0\rightarrow 1}$ was measured across the $0 \rightarrow 1$ electron transition.  A comparison of the curves in Fig.~\ref{fig:fig2}a shows that $g_{sens}$ drops as temperature rises immediately before the midpoint of the transition ($V_G-V_{mid}<0$), whereas $g_{sens}$ rises with temperature for $V_G-V_{mid}>0$.   Oscillating the reservoir temperature by $\delta T$ using an AC current through the heater while measuring oscillations in the charge sensor signal, $\delta g_{sens}$, with a lock-in amplifier (see Methods) yields the characteristic peak-dip structure seen in Fig.~\ref{fig:fig2}c. The expected lineshape of such a curve is given by $\delta g = \frac{\partial g}{\partial T} \delta T$, where $g(V;T)$ is defined by Eq.~\ref{eqn:g} and $\delta T$ is determined by the RMS value of $I_{heat}$:
%
\begin{align}
\label{eqn:dg}
        \delta g_{sens}(V_G, \Theta) &\propto -\delta T \left[ \frac{V_G-V_{mid}(\Theta)}{2\Theta} + \frac{\epsilon}{2} \right]\times \\
        				      &\quad\cosh^{-2}\left(\frac{V_G-V_{mid}(\Theta)}{2\Theta}\right) + const. \nonumber
\end{align}
%
where $\epsilon=\frac{\partial V_{mid}}{\partial \Theta}$. Using the lever arm to convert $V_{mid}$ to electrochemical potential, $\epsilon$ is simply the change in entropy (scaled by $k_B$) through Eq.~\ref{eqn:max}:
%
% 
%
\begin{align}
\label{eqn:epsilon}
        \epsilon &= \frac{\partial V_{mid}}{\partial \Theta} = 
        \frac{1}{k_B} \frac{\partial \mu}{\partial T} = 
        -\frac{1}{k_B} \Delta S_{N-1\rightarrow N}
\end{align}
%

From Eqs.~\ref{eqn:dg} and \ref{eqn:epsilon}, $\delta g_{sens}(V_G)$ would be perfectly antisymmetric around $V_{mid}$ if $\Delta S_{N-1\rightarrow N}$ were zero, whereas the data in Fig.~\ref{fig:fig2}c are clearly not antisymmetric. A best fit to Eqs.~\ref{eqn:dg} and \ref{eqn:epsilon} yields $\Delta S_{0\rightarrow 1}=(1.02 \pm 0.01) k_B \ln{2}$. For the $0 \rightarrow 1$ electron transition at zero magnetic field, one expects the following: $S_0$ for the empty state should be $k_B \ln{1}=0$ whereas for the spin degenerate 1-electron state one expects $S_1=k_B\ln{2}$, giving $\Delta S_{0\rightarrow 1}\equiv S_1 - S_0 =k_B\ln{2}$, matching the very well with the measured value.

Figure~\ref{fig:fig2}e shows that $\Delta S$, determined from fits analogous to those in Fig.~\ref{fig:fig2}c, remains constant over a range of $\delta T$ ($I_{heat}$ AC). The calibration for $\delta T$ with respect to $I_{heat}$ is shown in Figure~\ref{fig:fig2}d; a DC heater current was applied and fits to \label{eqn:g} yield $\delta T$ through $\Theta$; the calibration holds for AC currents when referring to the RMS value of $I_{heat}$ AC, as we do throughout this discussion. An important point is illustrated in Figure~\ref{fig:fig2}e; the change in entropy, $\Delta S$, is determined entirely by the fit parameter, $\epsilon$, with no further calibration necessary.

%%% figure 3 %%%
\begin{figure}
        \includegraphics[width=1.0\columnwidth]{../figures/figure_3.pdf}
        \caption{\label{fig:fig3}(a) Changes in entropy for $N=0 \rightarrow 1$ and $2 \rightarrow 3$ transitions, overlaid to highlight similar behaviors.  Each data point corresponds to a single $\delta g_{sens}(V_G)$ fit; multiple scans are carried out at various parallel magnetic fields.   (b) and (c) Characteristic $\delta g_{sens}$ traces from which the data in (a) were extracted. The two data points corresponding to (b) and (c) are shown as large markers in (a). (e) Bias spectroscopy data for the $N=0 \rightarrow 1$ transition. Dashed line at $V_{SD}$ = \SI{700}{\micro\electronvolt} shows where data in (f) are taken. (f) Fixed bias data showing fits to Zeeman splitting of the ground state (dashed lines) from which $g = -0.42 \pm 0.01$ is extracted.}
\end{figure}

Confirmation that the measured $\Delta S$ derives from spin degeneracy can be seen through its evolution with in-plane magnetic field, $B_\parallel$. Figure \ref{fig:fig3}a compares $\Delta S(B_\parallel)$ for the $0 \rightarrow1 $ and $2 \rightarrow 3$ electron transitions.  The 2-electron ground state is a spin singlet, while the 3-electron ground state adds an additional unpaired electron, so both $\Delta S_{0 \rightarrow 1}$ and $\Delta S_{2 \rightarrow 3}$ are expected to correspond to transitions from total spin zero to total spin one-half. The entropies of the 1- and 3-electron states are suppressed when Zeeman splitting lifts the spin degeneracy of the unpaired electron. This behavior is described by the Gibbs entropy for a two-level system:
%
\begin{align}
\label{eqn:gibbs}
        S &= k_B \sum_{i=\pm} p_{i}(B_\parallel, T) \ln{ p_{i}(B_\parallel,T) }
\end{align}
%
where $p_{\pm}(B_\parallel, T) = (1+ e^{\mp \frac{g\mu_B B_{\parallel}}{k_B T}})^{-1}$ are the probabilities for the unpaired electron to be in the spin up/down states at a given field and temperature. Fits to Eq.~\ref{eqn:gibbs} are shown in Fig.~\ref{fig:fig3}a for both transitions, with the ratio $g/T$ and an added scaling $\Delta S(B=0)$ as free parameters.

The entropy measurements yield $\Delta S_{0 \rightarrow 1}=(0.93 \pm 0.02) k_B \ln{2}$ and $\Delta S_{2 \rightarrow 3}=(0.98 \pm 0.01) k_B \ln{2}$ at zero magnetic field, collapsing to zero at high field where spin degeneracy is broken.  At a qualitative level, Figs.~\ref{fig:fig3}b and c show $\delta g_{sens}$ crossover from asymmetric to antisymmetric as $\Delta S$ goes to zero at high field. Choosing $T$ to be midway between the low ($T_{MC}$) and high ($T_{MC}+\delta T_{RMS}$) points of the temperature oscillation for each data set yields $g=-0.39 \pm 0.01 $ and $g=-0.45 \pm 0.01 $ for the $0\rightarrow 1$ and $2\rightarrow 3$ transitions, respectively. These are consistent with previous few electron dot measurements and the value measured separately in Fig.~\ref{fig:fig3}e using bias spectroscopy \cite{Cronenwett1998,Hanson2003,Zumbuhl2004}.

%%% figure 4 %%%
\begin{figure}
        \includegraphics[width=1.0\columnwidth]{../figures/figure_4.pdf}
        \caption{\label{fig:fig4}(a) Change in entropy, extracted from $\delta g_{sens}$ fits at varying parallel field. Dashed line shows fit to Eq.~\ref{eqn:gibbs}. Values stated for $\Delta S$ are with respect to the vertical offset apparent in the data. (b), (c), and (d) show characteristic $\delta g_{sens}$ traces from which the data in (a) were extracted. These data points are show as large markers in (a). (e) Bias spectroscopy data for the $N=1 \rightarrow 2$ transition. Transitions to the 2-electron triplet state correspond to the lines appearing at $V_{SD} = \SI{\pm 320}{\micro\electronvolt}$. Dashed line at $V_{SD}$ = \SI{1250}{\micro\electronvolt} shows where data in (f) are taken. (f) Fixed bias data in parallel field. Triplet level is split into $\mathit{T_+}$ and $\mathit{T_0}$ levels with a third $\mathit{T_-}$ level not visible here. At \SI[input-protect-tokens]{9}{\tesla} $\mathit{T_+}$ becomes degenerate with $\mathit{S}$. $g=-0.40\pm0.04$ is determined using $\mathit{T_0}$ and $\mathit{T_+}$ fits (dashed).}
\end{figure}

The $1\rightarrow 2$ transition (Fig.~\ref{fig:fig4}a) can be understood as the inverse of Fig \ref{fig:fig3}a for $B_\parallel < \SI{5}{\tesla}$.  There, the 2-electron ground state remains a spin singlet ($\mathit{S_2}=0$) while the 1-electron ground state entropy goes from $k_B\ln{2}$ to zero due to Zeeman splitting. At higher field, the 1-electron ground state remains non-degenerate ($\mathit{S_1}=0$) while the 2-electron ground state becomes two-fold degenerate where the spin singlet ($\mathit{S}$) and triplet ($\mathit{T_+}$) states cross. The singlet-triplet crossing is seen clearly around at \SI{9}{\tesla} in bias spectroscopy data for the $1\rightarrow 2$ transition (Fig.~\ref{fig:fig4}e).

The field-dependent entropy measurement for the 1-2 transition can again be fit using Eq.~\ref{eqn:gibbs}, with probabilities as before for the 1-electron states and $p_{S/T}(B_\parallel, T) = (1+ e^{\mp \frac{g\mu_B B_\parallel - \Delta_{ST}}{k_B T}})^{-1}$ for the 2-electron states, where $\Delta_{ST}$ is the singlet-triplet splitting at zero field. The fit allows for an offset from $\Delta S=0$, away from the degenerate points, to compensate for non-linearities in the charge sensor. From this we find $\Delta S$ at the two-fold degenerate points, $B=0$ and \SI{9}{\tesla}, are $-(1.01\pm0.02) k_B \ln{2}$ and $(1.043 \pm 0.02) k_B \ln{2}$, respectively. The extracted value of $g = 0.37 \pm 0.01$ from the low-field peak agrees well with the expected value and the bias spectroscopy data in Fig.~\ref{fig:fig4}e. At the high-field singlet-triplet degeneracy we find $g = 0.55 \pm 0.02$. This unexpectedly high g-factor can be explained by a shifting of the $\mathit{T_0}$ state with magnetic field, as seen in Fig.~\ref{fig:fig4}f and previous work \cite{Szafran2004}. The high quality of the Gibbs entropy fit confirms $\Delta S$ and illustrates a tunable degeneracy in the system.

%%% conclusion %%%
In conclusion, we show that the entropy of few-particle states in mesoscopic devices can be measured directly.  Looking at the well-understood spin degeneracy of  few-electron ground states states in a GaAs quantum dot for calibration, the entropy can be quantified within 5\% of the expected $\pm k_B\ln{2}$.  Future work may take advantage of this strategy to search for thermodynamic signatures of non-trival electron and quasiparticle states, eliminating the uncertainty found in typical transport data.
